Behavior Driven Development in Haskell

### 2012-06-01 paper.tex
Add URL for testpack

This section show best current practices of
BDD in Haskell.

\subsection{Procedure of BDD}

First, we write a signature and leave the definition of the function undefined.

\begin{quote}
\small
\begin{verbatim}
reverse :: [a] -> [a]
reverse = undefined
\end{verbatim}
\end{quote}

\noindent Next we write user documentation in Haddock markup:

\begin{quote}
\small
\begin{alltt}
-- |
-- 'reverse' @xs@ returns the elements of @xs@
-- in reverse order. @xs@ must be finite.
--
-- >>> reverse [1,2,3]
-- [3,2,1]
--
-- prop> reverse [] == reverse []
-- prop> reverse (xs ++ ys) == reverse ys ++ \(\hookleftarrow\)
                               reverse (xs::[Int])
-- prop> reverse (reverse xs) == (xs::[Int])

reverse :: [a] -> [a]
reverse = undefined
\end{alltt}
\end{quote}

\noindent We should execute Doctest to ensure no syntax errors.
Also, we write code documentation with Hspec:

\begin{quote}
    sol: TBD. @kazu, the descriptions make no sense, can you try to
    make them more precise?
\end{quote}

\begin{quote}
\small
\begin{verbatim}
describe "reverse" $ do
  it "returns the first element in the last" $
    property $ \xs -> not (null xs) ==>
      head (reverse xs) == last (xs :: [Int])

  it "returns the last element in the first" $
    property $ \xs -> not (null xs) ==>
      last (reverse xs) == head (xs :: [Int])
\end{verbatim}
\end{quote}

\noindent Also, we should execute this Hspec module
to ensure no syntax errors.
Then we implement the function.

\begin{quote}
\small
\begin{verbatim}
reverse :: [a] -> [a]
reverse = foldl (flip (:)) []
\end{verbatim}
\end{quote}

\noindent Now we run tests, and improve our implementation, until all
tests pass.

@@

The {\tt testpack} package provides a combinator to expect exceptions,
similar to {\tt shouldThrow} from
Section~\ref{sec:expacting-exceptions}:

>>>>

This section shows how behaviour driven development works in practice,
by means of the fibonacci function.  Of course this is a very simple
example, but the method is also applicable to more sophisticated
situations where design decisions might not be as obvious as here.

We present a step-by-step refinement of a single file {\tt Fib.hs}
that contains both implementation and specs.  Each line of code is
only given once when introduced, subsequent code fragments are
incremental updates.  The reader is asked to substitute code parts as
appropriate.

\subsection{Step 1}

We start with the function's documentation, which gives a usage
example, together with the simplest possible implementation that makes
the example work.

\begin{quote}
\small
\begin{verbatim}
module Fib where
-- | Calculate Fibonacci numbers.
--
-- >>> fib 10
-- 55
fib :: Int -> Integer
fib = const 55
\end{verbatim}
\end{quote}

This first step might appear useless, but as of now we already have
made some choices with regard to the API of our library.  By giving a
usage example, we were forced to think about which arguments our
function takes.  We can now run Doctest to verify that the example
works.

\begin{quote}
\small
\begin{verbatim}
$ doctest Fib.hs
There is one test, with one single interaction.
Examples: 1  Tried: 1  Errors: 0  Failures: 0
\end{verbatim}
\end{quote}

\subsection{Step 2}

We add a specification that forces our function to actually calculate
fibonacci numbers.

\begin{quote}
\small
\begin{verbatim}
import Control.Applicative
import Test.Hspec.ShouldBe
import Test.QuickCheck

-- small non-negative numbers
newtype Small = Small Int
  deriving Show

instance Arbitrary Small where
  arbitrary = Small . (`mod` 10) <$> arbitrary

main = hspec $ do
  describe "fib" $ do
    it "calculates arbitrary Fibonacci numbers" $ do
      property $ \(Small n) ->
        fib n == fib (n + 2) - fib (n + 1)
\end{verbatim}
\end{quote}

\noindent Of course, this test now fails when we run it,

\begin{quote}
\small
\begin{verbatim}
$ runhaskell Fib.hs
fib
 - calculates arbitrary Fibonacci numbers FAILED [1]
\end{verbatim}
\end{quote}

\noindent so we add a working implementation.  Both Doctest and Hspec
are happy now.

\begin{quote}
\small
\begin{verbatim}
fib 0 = 0
fib 1 = 1
fib n = fib (n - 1) + fib (n - 2)
\end{verbatim}
\end{quote}

\subsection{Step 3}

As the seasoned Haskeller might have already noticed, our
implementation is quite inefficient, as can be proved by this
specification:

\begin{quote}
\small
\begin{verbatim}
import Control.Exception
import System.Timeout

it "is efficient" $ do
  timeout 10000
      (evaluate $ fib 32) `shouldReturn` Just 2178309
\end{verbatim}
\end{quote}

\noindent When we run it, hspec tells us

\begin{quote}
\small
\begin{verbatim}
$ runhaskell Fib.hs
fib
 - calculates arbitrary Fibonacci numbers
 - is efficient FAILED [1]
\end{verbatim}
\end{quote}

\noindent and we respond with an efficient implementation.

\begin{quote}
\small
\begin{verbatim}
import Data.List

fib :: Int -> Integer
fib n = fibs !! n
fibs = 0 : 1 : zipWith (+) fibs (tail fibs)
\end{verbatim}
\end{quote}

\subsection{Step 4}

So far, so good.  We have an implementaion of the fibonacci function
that swiftly calculates correct values for all positive integers.
But what about negative ones? Our implementation uses the list index
operator {\tt (!!)} which is a partial function that throws an
exeption for negative indexes.  Let's specify that behaviour.

\begin{quote}
\small
\begin{verbatim}
it "throws ErrorCall on negative input" $ do
  evaluate (fib (-10)) `shouldThrow` anyErrorCall
\end{verbatim}
\end{quote}

\noindent Throwing exceptions from pure code is considered bad
practice, so we guard the {\tt (!!)} operator against negative
indexes.

\begin{quote}
\small
\begin{verbatim}
fib n | n < 0     = 0
      | otherwise = fibs !! n
\end{verbatim}
\end{quote}

\noindent Now our specification fails, so we adapt it accordingly.

\begin{quote}
\small
\begin{verbatim}
it "returns 0 on negative input" $ do
  fib (-10) `shouldBe` 0
\end{verbatim}
\end{quote}

\noindent In fact, it is the case that fib returns 0 for all negative
numbers.

\begin{quote}
\small
\begin{verbatim}
it "returns 0 on negative input" $ do
  property $ \n -> n < 0 ==>
    fib n == 0
\end{verbatim}
\end{quote}

\noindent This concludes our example coding session.  We have seen how
to utilize behavior driven development to incrementally build a small
library.  The combinators {\tt shouldBe}, {\tt shouldReturn} and {\tt
shouldThrow} can be used to specify behavior of functions in an
intuitive, easy to read manner.  The final version of our library
looks like this.

\begin{quote}
\small
\begin{verbatim}
module Fib where
import Control.Applicative
import Control.Exception
import Data.List
import System.Timeout
import Test.Hspec.ShouldBe
import Test.QuickCheck

-- | Calculate Fibonacci numbers.
--
-- >>> fib 10
-- 55
fib :: Int -> Integer
fib n | n < 0     = 0
      | otherwise = fibs !! n

fibs = 0 : 1 : zipWith (+) fibs (tail fibs)

-- small non-negative numbers
newtype Small = Small Int
  deriving Show

instance Arbitrary Small where
  arbitrary = Small . (`mod` 1000) <$> arbitrary

main = hspec $ do
  describe "fib" $ do
    it "calculates arbitrary Fibonacci numbers" $ do
      property $ \(Small n) ->
        fib n == fib (n + 2) - fib (n + 1)

    it "is efficient" $ do
      timeout 10000 (evaluate $ fib 32)
          `shouldReturn` Just 2178309

    it "returns 0 on negative input" $ do
      property $ \n -> n < 0 ==>
        fib n == 0
\end{verbatim}
\end{quote}

@@

The {\tt testpack}\footnote{%
http://hackage.haskell.org/package/testpack}
package provides a combinator to expect exceptions,
similar to {\tt shouldThrow} from
Section~\ref{sec:expacting-exceptions}:
